Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Cache chunks manuscript

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf heckman models
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

@@

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred["popular", "school"] <- 1 
plot_pred(pred)

@@

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

@@

```{r pop_multilevel, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred["popular", "school"] <- -2 
plot_pred(pred)

@@

Impute the incomplete data
```{r imp_impact}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

```{r impact-fit}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

Select relevant variables:
```{r bmi}
dat <- select(obesity, -"bmi")
```

@@

Create predictor matrix:
```{r obesity-pred}
pred <- quickpred(dat)   # predictor matrix
pred["weight","cluster"]<- -2 # clustering variable
pred["weight","rt"] <- -3 #inclusion-restriction variable
plot_pred(pred)
```

@@

Impute the missingness:
```{r eval=FALSE}
imp <- mice(
  dat, # dataset with missing values
  m = 5, # number of imputations
  maxit = 10,
  meth = meth, #imputation method vector
  pred = pred, #imputation predictors matrix
  meta_method = "reml",
  printFlag = FALSE
)
```

>>>>

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf heckman models
library(metamisc)     # for ???
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

@@

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

@@

```{r pop_predictor, message=FALSE, warning=FALSE, cache=TRUE}
# adjust the predictor matrix
pred["popular", "school"] <- 1 
plot_pred(pred)

@@

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

@@

```{r pop_multilevel, message=FALSE, warning=FALSE, cache=TRUE}
# adjust the predictor matrix
pred["popular", "school"] <- -2 
plot_pred(pred)

@@

Impute the incomplete data
```{r imp_impact, cache=TRUE}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

Select relevant variables:
```{r bmi}
dat <- obesity
```

@@

Create predictor matrix:
```{r obesity-pred}
pred <- quickpred(dat)   # predictor matrix
pred["Weight","Cluster"]<- -2 # clustering variable
pred["Weight", ] <- -3 #inclusion-restriction variable
plot_pred(pred)
```

@@

Impute the missingness:
```{r eval=FALSE, cache=TRUE}
imp <- mice(
  dat, # dataset with missing values
  m = 5, # number of imputations
  maxit = 10,
  meth = meth, #imputation method vector
  pred = pred, #imputation predictors matrix
  meta_method = "reml",
  printFlag = FALSE
)
```

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Cache chunks manuscript

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf heckman models
\end{CodeInput}
\end{CodeChunk}

@@

In this section we'll go over the different steps involved with imputing
incomplete multilevel data with the R package mice. We consider the
simulated \texttt{popmis} dataset, which included pupils (\(n = 2000\))
clustered within schools (\(N = 100\)). The following variables are of
primary interest:

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- select(obesity, -"bmi")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred["weight","cluster"]<- -2 # clustering variable
R> pred["weight","rt"] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

@@

\begin{itemize}
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\end{itemize}

>>>>

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf heckman models
R> library(metamisc)     # for ???
\end{CodeInput}
\end{CodeChunk}

@@

In this section we will go over the different steps involved with
imputing incomplete multilevel data with the R package mice. We consider
the simulated \texttt{popmis} dataset, which included pupils
(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
variables are of primary interest:

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1}

\caption[True effect in the popularity data]{True effect in the popularity data}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- obesity
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred["Weight","Cluster"]<- -2 # clustering variable
R> pred["Weight", ] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

@@

\begin{itemize}
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\item
  FIML vs MI
\end{itemize}

An alternative approach to missing data is to use Full Information
Maximum Likelihood (FIML). This method does not require the imputation
of any missing values. Whereas MI consists of imputation, analyses and
pooling steps, FIML analyses the data in a single step. When the
assumptions are met the two approaches should produce equivalent
results. {[}REF{]} As FIML requires specialised software, not all
analyses can be performed with standard software. {[}REF{]}

\begin{itemize}
\tightlist
\item
  Survival / TTE, this could be put in the paragraph on congeniality
\end{itemize}

When the outcome is time-to-event, the Nelson-Aalen estimate of the time
to event should be included as a covariate in the imputation model
{[}REF{]}

In hierarchical datasets, clustering is a concern because the
homoscedasticity in the error terms cannot be assumed across clusters
and the relationship among variables may vary at different hierarchical
levels. When multiple imputation is used to deal with missing data, as
the imputation and analysis process is performed separately, it is
necessary that imputation model being congenial with the main analysis
model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
structure also imputation model should do it (Audigier, 2021). Not
including clustering into the imputation process may lead to effect
estimates with smaller standard errors and inflated type I error.

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})

The selection of each strategy depends mainly on the assumptions in the
main analysis and also on the restriction of the analyzed data.

Regarding the restrictions imposed by the data, for instance, the use of
cluster indicator variables is restricted in datasets where there are
not many clusters and many observations per cluster (Graham, 2009). The
last restriction is also required when imputations are performed on each
cluster separately. When this restriction cannot be achieved, one can
use an imputation model that simultaneously imputes all clusters using a
hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modelled by a random effect so
the hierarchical model can be estimated even when there are few
observations per cluster. However, this strategy is best suited for
balanced data (Grund, 2017) and when random effects model is
appropriated, i.e.~the number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance by using the cluster indicator strategy may lead to
bias estimates when the model is based on a hierarchical model
(Taaljard,2008). Even when an imputation strategy congenial with the
main model is preferred, it is important to consider whether it is
appropriate for the data as a less complex imputation strategies may
also lead to unbiased estimates in certain scenarios(Bailey 2020). For
instance, in causal effect analysis, separately imputation may lead to
smaller bias when the size of the smaller exposure cluster is large,
compared with an imputation model that includes exposure-confounder
interactions. (Zhang,2023).
